---
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  rmd: ""
  title: "EDA: General (Numeric Matrix)"
  digits: 2
title: "`r params$title`"
output:
  html_document:
    theme: journal
    highlight: textmate
    df_print: kable
    toc: true
    toc_depth: 2
    code_folding: hide 
---

```{r load_knitr, include = FALSE}
library(knitr)
opts_chunk$set(message = FALSE)
```

```{r setup}
library(janitor)
library(tidyverse)

dat <- as.matrix(read.csv(snakemake@input[[1]], row.names = 1))
```

```{r helper_function}
# helper function to round numbers, including trailing zeros
round_ <- function(x) {
    formatC(round(x, params$digits), params$digits, format = "f")
}
```

## Dimensions

- \# rows: `r nrow(dat)`
- \# cols: `r ncol(dat)`

## Distribution

```{r}
qplot(as.numeric(as.matrix(dat)), geom = 'density', fill = I('blue'), alpha = 0.75) +
  xlab("Value") + 
  theme_bw() +
  theme(legend.position = 'none')
```

- min: `r round_(min(dat, na.rm = TRUE))`
- mean: `r round_(median(dat, na.rm = TRUE))`
- median: `r round_(median(dat, na.rm = TRUE))`
- max: `r round_(max(dat, na.rm = TRUE))`
- quartiles: `r round_(quantile(dat, na.rm = TRUE))`

## Values

- \# missing values: `r sum(is.na(dat))`
- \# zeros: `r sum(dat == 0, na.rm = TRUE)`
- \# positive values: `r sum(dat > 0, na.rm = TRUE)`
- \# negative values: `r sum(dat < 0, na.rm = TRUE)`

## Columns

**Column types**:

```{r column_types}
type <- apply(dat, 2, class)
tabyl(type)
```

**Column statistics**:

```{r}
col_means    <- round_(range(apply(dat, 2, mean, na.rm = TRUE)))
col_medians  <- round_(range(apply(dat, 2, median, na.rm = TRUE)))
col_std_devs <- round_(range(apply(dat, 2, sd, na.rm = TRUE)))
```

- Column mean (min): `r col_means[1]`
- Column mean (max): `r col_means[2]`
- Column median (min): `r col_medians[1]`
- Column median (max): `r col_medians[2]`
- Column stdev (min): `r col_std_devs[1]`
- Column stdev (max): `r col_std_devs[2]`

## Rows

**Row statistics**:

```{r}
row_means    <- round_(range(apply(dat, 2, mean, na.rm = TRUE)))
row_medians  <- round_(range(apply(dat, 2, median, na.rm = TRUE)))
row_std_devs <- round_(range(apply(dat, 2, sd, na.rm = TRUE)))
#row_outliers <- self$detect_row_outliers()

# clip outlier lists if more than a few
#if (length(row_outliers) > 5) {
#    row_outliers <- c(head(row_outliers, 5), '...')
#}
#if (length(col_outliers) > 5) {
#    col_outliers <- c(head(col_outliers, 5), '...')
#}
```

- Row mean (min): `r row_means[1]`
- Row mean (max): `r row_means[2]`
- Row median (min): `r row_medians[1]`
- Row median (max): `r row_medians[2]`
- Row stdev (min): `r row_std_devs[1]`
- Row stdev (max): `r row_std_devs[2]`

## Outliers

### Column outliers

```{r, results = 'asis'}
# compute column pairwise correlation matrix
column_cor_mat <- cor(dat)

# detect column outliers (> 2 stdevs away)
median_column_cors <- apply(column_cor_mat, 2, median)
cutoff <- mean(median_column_cors) - (2 * sd(median_column_cors))

col_outliers <- colnames(dat)[median_column_cors < cutoff]

if (length(col_outliers) > 0) {
  cat(paste0('- ', str_flatten(col_outliers, '\n- '), '\n'))
} else {
  cat('- None\n')
}
```

### Row outliers

```{r, results = 'asis'}
# compute row pairwise correlation matrix
row_cor_mat <- cor(t(dat))

# detect row outliers (> 2 stdevs away)
median_row_cors <- apply(row_cor_mat, 1, median)
cutoff <- mean(median_row_cors) - (2 * sd(median_row_cors))

row_outliers <- colnames(dat)[median_row_cors < cutoff]

if (length(row_outliers) > 0) {
  cat(paste0('- ', str_flatten(row_outliers, '\n- '), '\n'))
} else {
  cat('- None\n')
}
```


